### 2019-02-12 paper/birth_spacing_lit.tex
Add notes on papers

(This might be something to test in a new paper. Mortality risk should
go down differentially across parities with access to sex selection,
depending on the sex of the prior child. This effect should be larger
for families more likely to use sex selection, i.e. higher education
women.
Since we do not know actual use of sex selection, one possibility
would be to used the area's sex ratio for the type of family.
A caveat is that sibling competition is less likely to occur, which
would also reduce mortality for the prior child, although possibly
less.
A first step would be to check on mortality levels among the
well-educated women.
)

@@

>>>>

@@

we use the estimates to perform simulations that show how the use of
contraceptives is linked to the infant mortality rate in two ways: (1)
by reducing mortality among higher order births by lengthening birth
intervals, and (2) through a composition effect: reducing the number of
higher order births leads to a larger proportion of relatively
vulnerable first order births, thereby increasing the average mortality
rate for all birth orders.

keeping other factors constant (including the decision to use
contraceptives at any time after the previous birth), birth intervals
tend to be shorter for higher birth orders, which is consistent with the
association between short birth intervals and high fertility.

\citep{Jayachandran2011}

If son preference exists, girls will be weaned earlier than boys because
parents want to try again for a son. Given the large health benefits of
breastfeeding in environments with contaminated food and water (Feachem
and Koblinsky 1984; Palloni and Millman 1986; Habicht, DaVanzo, and Butz
1988), such behaviorcanleadtogenderdisparities inchildhealth. Incontrast
to the standard explanation of gender health disparities—that parents
actively invest more in sons’ health—we show that if parents merely
prefer to have sons, then mothers will breastfeed daughters less,
causing a gender gapin health even when parents
equallyvaluethehealthofall theirchildren. Ourresults highlight how
health disparities can arise “passively” from fertility preferences, and
relate to earlier work (Yamaguchi 1989; Clark 2000; Jensen 2003) showing
that a “try until you have a son” fertility rule results in girls, on
average, having more siblings and thus more competition for household
resources.

Model predictions:
First, breastfeeding duration increases with birth order, as demand for
breastfeeding’s contraceptive properties should rise as couples approach
or exceed their ideal family size, andin fact shouldincrease
discontinuously once ideal family size is met.
Second, as explained, as long as some son preference exists, girls are
breastfed less than boys.
Third, by the same logic, children with older brothers are breastfed more.
Fourth, the gender effect is largest when a couple approaches its target
family size since, at that point, their decision to have another child
is highly marginal and thus most sensitive to sex composition; the
effect is smallest for children whose birth order is either
significantly below or above ideal family size because couples will want
to, respectively, wean or nurse such children regardless of gender.

several papers show empirically that in settings with son preference, a
couple that just had a son is more likely to stop having children (Das
1987) or wait longer to have the next child (Rindfuss et al. 1982;
Trussell et al. 1985; Arnold, Choe, and Roy 1998; RetherfordandRoy
2003). Hemochandra, Singh, and Singh (2010) show that birth intervals
are shorter if parents have not reached their self-reported ideal number
of sons. Second, Nemeth and Bowling (1985) examine whether having nosons
shortens the duration of breastfeeding.

Muhuri and Menken (1997) show that birth spacing and the sex of older
siblings affect mortality and state that shorter birth intervals after
girls are one potential cause of excess female mortality. In unreported
results, Arnold, Choe, and Roy (1998) examine whether the gender gap in
mortality in India is due to birth spacing; they donot findevidence
tosupport the hypothesis. Finally, Mutharayappa et al. (1997) show that
there is a gender gap in breastfeeding in India and conjecture that it
is related to son-biased fertility patterns.

((Bhatia 1978; Basu 1992) on lower son preference in the south)

Retherford et al. (1989) find that controlling for breastfeeding largely
eliminates the negative correlation between infant mortality and
subsequent birth spacing in Nepal.

\citep{Whitworth2002}

This paper examines the impact of the length of the preceding birth
interval on under-two mortality in India, and examines the pathways
through which short preceding birth intervals may lead to an increased
risk of mortality

1992 NFHS

Three mortality periods are examined: neonatal, early post neonatal and
late post-neonatal and toddler

short preceding birth intervals (< 18 months) are associated with an
increased risk of mortality in all three age groups, and the effect is
particularly marked in the early post-neonatal period.

diluting effect that a higher level of maternal education has on the
relationship between short preceding birth intervals and mortality risk

There is evidence to suggest that sibling rivalry is a pathway through
which short birth intervals inﬂuence mortality, with the death of the
previous sibling removing the competition for scarce resources, and
resulting in lower risks of mortality than if the previous sibling was
still alive.

\citep{Bhalotra2008}
